---
title: "Early EDA"
output: html_notebook
---

Packages
```{r}
library(tidyverse)
library(ggplot2)
library(scatterplot3d)
#library(rgl) I can't download this package but it is supposed to make the 3d scatterplot interactive
```

Data
```{r}
at_bat <- read.csv("C:/Users/lucas/OneDrive - The Pennsylvania State University/DS 440/DS440/at_bat.csv")
head(at_bat)
```
```{r}
#grouping the out categories
at_bat <- at_bat %>%
  mutate(events_gr = case_when(
    events %in% c("fielders_choice_out", "force_out", "grounded_into_double_play",
                  "sac_fly", "fielders_choice", "triple_play", "double_play",
                  "sac_fly_double_play") ~ "field_out",
    TRUE ~ events
  ))

unique(at_bat$events_gr)
unique(at_bat$events)
```

Univariate
```{r}
cat("Launch Angle 5 Number Summary\n")
summary(at_bat$launch_angle)

cat("\nExit Velocity 5 Number Summary\n")
summary(at_bat$launch_speed)
```

```{r}
hist(at_bat$launch_angle, 
     main = "Histogram of Launch Angle",
     xlab = "Launch Angle",
     ylab = "Frequency"
)

hist(at_bat$launch_speed, 
     main = "Histogram of Exit Velocity",
     xlab = "Exit Velocity",
     ylab = "Frequency"
)
```
Launch angle is pretty normal which you would come to expect seeing that all players want to hit the ball coming across it or that is what they are aiming to do. Exit velocity is left skewed with a median of 91, which also makes sense as players do need to hit the ball hard and well to have a ball in play. 

```{r}
#spray angle percentage:
at_bat <- at_bat %>%
  mutate(pull = case_when(
    stand == "R" & spray_angle < -20 ~ "pulled",
    stand == "R" & spray_angle >= -20 & spray_angle <= 20 ~ "center",
    stand == "R" & spray_angle > 20 ~ "opposite",
    stand == "L" & spray_angle < -20 ~ "opposite",
    stand == "L" & spray_angle >= -20 & spray_angle <= 20 ~ "center",
    stand == "L" & spray_angle > 20 ~ "pulled",
    TRUE ~ NA_character_  # Handle other cases if any
  ))

pulled_percentages <- at_bat %>%
  group_by(stand) %>%
  summarise(pulled_percentage = sum(pull == "pulled") / n() * 100)

print(pulled_percentages)

events_of_interest <- c("field_out", "single", "double", "home_run", "triple")
filtered_data <- at_bat %>%
  filter(events_gr %in% events_of_interest)

ggplot(filtered_data, aes(x = pull, fill = events_gr)) +
  geom_bar(position = "fill", stat = "count") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(title = "Percentage of Events by Pull",
       x = "Pull",
       y = "Percentage") +
  scale_fill_manual(values = c("field_out" = "blue", 
                               "single" = "green", 
                               "double" = "orange", 
                               "home_run" = "red", 
                               "triple" = "purple"))
```



Bivariate
```{r}
#only out, single, double, triple, HR
main_events <- c("field_out", "single", "double", "triple", "home_run")

custom_colors <- c(field_out = "blue", single = "yellow", double = "orange", triple = "green", home_run = "red")

subset_data <- subset(at_bat, events %in% main_events)

plot(subset_data$launch_speed, 
     subset_data$launch_angle, 
     col = custom_colors[subset_data$events],
     main = "Scatterplot of Launch Speed vs Spray Angle With Outs",
     xlab = "Launch Speed",
     ylab = "Spray Angle"
)

legend("topright",
       legend = names(custom_colors), 
       col = custom_colors, 
       title = "Event",
       pch = 16,
       cex = 0.8
)
```

```{r}

scatterplot3d(subset_data$launch_speed, 
              subset_data$launch_angle, 
              subset_data$spray_angle, 
              color = custom_colors[subset_data$events],
              main = "3D Scatterplot of Launch Speed, Launch Angle, and Spray Angle",
              xlab = "Launch Speed",
              ylab = "Launch Angle",
              zlab = "Spray Angle"
)


legend("topright",
       legend = names(custom_colors), 
       fill = custom_colors,
       title = "Event",
       cex = 0.8
)

rglwidget()
```

```{r}
#only single, double, triple, HR
main_events <- c("single", "double", "triple", "home_run")

subset_data <- subset(at_bat, events %in% main_events)

custom_colors <- c("red", "yellow", "green", "blue")

plot(subset_data$launch_speed, 
     subset_data$spray_angle, 
     col = custom_colors[as.factor(subset_data$events)],
     main = "Scatterplot of Launch Speed vs Spray Angle Without Outs",
     xlab = "Launch Speed",
     ylab = "Spray Angle"
)

legend("topright",
       legend = unique(subset_data$events), 
       col = custom_colors, 
       title = "Event",
       pch = 16,
       cex = 0.8
)
```
You can see a lot of the homeruns are pulled when they are weaker hit balls which I think is interesting to note. With regular xwOBA only accounting for launch speed (exit velo) it probably wouldn't count this very well. 

```{r}
#only single, double, triple, HR
main_events <- c("single", "double", "triple", "field_out")

subset_data <- subset(at_bat, events %in% main_events)

custom_colors <- c("red", "yellow", "green", "blue")

plot(subset_data$launch_speed, 
     subset_data$sprint_speed, 
     col = custom_colors[as.factor(subset_data$events)],
     main = "Scatterplot of Launch Speed vs Sprint Speed",
     xlab = "Launch Speed",
     ylab = "Sprint Speed"
)

legend("topright",
       legend = unique(subset_data$events), 
       col = custom_colors, 
       title = "Event",
       pch = 16,
       cex = 0.8
)
```

```{r}

```

